preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-07 Manuscript/manuscript.Rmd
First shot on Abstract

# ABSTRACT (Spiro version)

**Background** Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associated Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by conventional TAD calling algorithms.

**Methods** To aid in the precise identification of TAD boundaries we developed preciseTAD, a random forest based framework that leverages the spatial relationship of high resolution ChIP-seq defined genomic elements, coupled with density-based clustering and scalable partitioning techniques. We benchmarked our method against a popular TAD-caller and a novel chromatin loop prediction algorithm on multiple cell lines.

**Results** We show that our framework performs well when predicting both TAD and chromatin loop boundaries. Compared to established techniques, boundaries predicted by preciseTAD are more enriched for CTCF, RAD21, SMC3, and ZNF143 sites. Likewise, preciseTAD boundaries were more conserved across cell lines, highlighting their biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using 1D genome annotation data. The precise identification of TAD boundaries will improve our understanding of how epigenetics shapes the 3D structure of the genome.

>>>>

# ABSTRACT

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, typically mapped at higher precision. Several methods have been developed to leverage genome annotation  data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods** To improve the detection of domain boundaries, we developed preciseTAD, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to investigate class imbalance. We benchmarked our method against Arrowhead TAD caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

**Results** Using different feature engineering and resampling strategies, we found spatial relationships between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for nearby CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. Our method is implemented as a preciseTAD R package (Bioconductor URL). The precise identification of TAD boundaries will improve our understanding of how epigenomics shapes the 3D structure of the genome.
